Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-08-21 Manuscript/ms.tex

	\begin{abstract}
		Graph neural network has become a research hotspot in the field of artificial intelligence due to state-of-the-art of performance
		achieved in many graph-related tasks. At the same time, various of graph neural network systems have emerged one after another.
		These systems have used their own skills in implementation, but there are very few works to study the performance bottleneck
		on graph neural network training. In contrast to prior works that present characterization of GCNs and select GNNs cover GNN workloads,
		we select typical algorithms by complexities of edge/vertex calculation. We analyze the performance bottleneck by breaking down training
		time, study the affects of hyper-parameters and sampling technologies and find factors of memory usage. We finally put forward some key
		findings for GNN training and hope our work can help GNN system researchers.
	\end{abstract}

>>>>

	\begin{abstract}
        Graph neural network(GNN) has become a research hotspot in the field of artificial intelligence due to state-of-the-art performance achieved in many graph-related tasks. At the same time, various graph neural network systems have emerged one after another.
        These systems have used their skills in implementation, but there are very few works to study the performance bottleneck on graph neural network training. In contrast to prior works that present characterization of GCNs and select GNNs cover GNN workloads,
		we select typical algorithms by complexities of edge/vertex calculation. We analyze the performance bottleneck by breaking down training time, study the effects of hyper-parameters and sampling technologies, and find factors of memory usage. 
		We finally put forward some key findings for GNN training and hope our work can help GNN system researchers.
	\end{abstract}

### 2020-08-21 Manuscript/sec2-review-of-graph-neural-network.tex

$\phi^l$ is the \emph{message} function.
For every incident edge $(v_j, v_i)$ of $v_i$, $\phi$ receives the output hidden feature vectors $\boldsymbol{h}^l_i$ and $\boldsymbol{h}^l_j$ of the previous GNN layer and the edge feature vector $\boldsymbol{e}_{j,i}$ as the input.
$\phi^l$ emits a message vector $\boldsymbol{m}^l_{j,i}$ for every edge $(v_j, v_i)$ at layer $l$, i.e., $\boldsymbol{m}^l_{j,i}=\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j, \boldsymbol{e}_{j,i})$.
For $v_i$, the message vectors $\boldsymbol{m}^l_{x,j}$ of its adjacent edges ($v_x \in \mathcal{N}(v_i)$) are aggregated by the \emph{aggregation} function $\Sigma^l$ to produce an aggregated vector $\boldsymbol{s}^l_i$, i.e., $\boldsymbol{s}^l_i=\mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}\boldsymbol{m}^l_{j,i}$.
$v_i$'s aggregated vector $\boldsymbol{s}^l_i$ and its hidden vector $\boldsymbol{h}^l_i$ from the previous GNN layer are fed into the \emph{vertex update} function $\gamma^l$ to calculate the output hidden vector $\boldsymbol{h}^{l+1}_i$ of the current layer $l$, i.e., $\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \boldsymbol{s}^l_i)$
The end-to-end training requires $\phi^l$ and $\gamma^l$ (like multi-layer perceptrons and GRU) and $\Sigma_l$ (like mean, sum, element-wise min/max) are \emph{differentiable} to make the whole GNN differentialble.

>>>>

$\phi^l$ is the \emph{message} function.
For every incident edge $(v_j, v_i)$ of $v_i$, $\phi$ receives the output hidden feature vectors $\boldsymbol{h}^l_i$ and $\boldsymbol{h}^l_j$ of the previous GNN layer and the edge feature vector $\boldsymbol{e}_{j,i}$ as the input.
$\phi^l$ emits a message vector $\boldsymbol{m}^l_{j,i}$ for every edge $(v_j, v_i)$ at layer $l$, i.e., $\boldsymbol{m}^l_{j,i}=\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j, \boldsymbol{e}_{j,i})$.
For $v_i$, the message vectors $\boldsymbol{m}^l_{x,j}$ of its adjacent edges ($v_x \in \mathcal{N}(v_i)$) are aggregated by the \emph{aggregation} function $\Sigma^l$ to produce an aggregated vector $\boldsymbol{s}^l_i$, i.e., $\boldsymbol{s}^l_i=\mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}\boldsymbol{m}^l_{j,i}$.
$v_i$'s aggregated vector $\boldsymbol{s}^l_i$ and its hidden vector $\boldsymbol{h}^l_i$ from the previous GNN layer are fed into the \emph{vertex update} function $\gamma^l$ to calculate the output hidden vector $\boldsymbol{h}^{l+1}_i$ of the current layer $l$, i.e., $\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \boldsymbol{s}^l_i)$
The end-to-end training requires $\phi^l$ and $\gamma^l$ (like multi-layer perceptrons and GRU) and $\Sigma_l$ (like mean, sum, element-wise min/max) are \emph{differentiable} to make the whole GNN differentiable.

### 2020-08-21 Manuscript/sec4-experiments.tex

\figurename~\ref{fig:exp_vertex_edge_cal_proportion} decomposes the training time of a GNN on the layer level.
The training time of each layer is the summation of the time in the forward, backward, and evaluation phases.
In GCN, GAT, and GaAN, the time spent on layer 0 was much larger than layer 1.
In those GNNs, the dimensions of the input/output feature vectors in the layer 0 were much larger than the dimensions in the layer 1.
$d^0_{in}=dim(\boldsymbol{x})$, $d^0_{out}=d^1_{in}=64$ and $d^1_{out}=\#Class$ and $dim(\boldsymbol{x}) \gg \#Class$.
For GaAN, since it required the dimensions of the input/output feature vectors must be the same, $d^0_{in}=d^0_{out}=d^1_{in}=d^1_{out}=64$ and the training time of both layers were close.

@@

In general, the most time-consuming operator in GNN calculation is still the matrix multiplication \texttt{mm} and the elementwise multiplication \texttt{mul}, \emph{making GNN training suitable for GPU}.
Although the aggregate step in the edge calculation is relatively simple (like sum and mean), the related operators \texttt{scatter\_add} and \texttt{gather} still consumed a certain amount of the time.
They have to synchronize between hardware threads to avoid updating the same aggregated vector at the same time and they conducted non-regular memory access with the access pattern determined by the edge set dynamically.
For GPU, they were less efficient than \texttt{mm}.
The index-based selection \texttt{index\_select} operator from the collection step consumed around 10\% of time in all GNNs.
Though GPUs have high on-chip memory bandwidth, improving the efficiency of \texttt{scatter\_add}/\texttt{gather}/\texttt{index\_select} (like overlapping them with $\phi$) can benifit the training of all kinds of GNNs.

>>>>

\figurename~\ref{fig:exp_vertex_edge_cal_proportion} decomposes the training time of a GNN on the layer level.
The training time of each layer is the summation of the time in the forward, backward, and evaluation phases.
In GCN, GAT, and GaAN, the time spent on layer 0 was much larger than layer 1.
In those GNNs, the dimensions of the input/output feature vectors in layer 0 were much larger than the dimensions in layer 1.
$d^0_{in}=dim(\boldsymbol{x})$, $d^0_{out}=d^1_{in}=64$ and $d^1_{out}=\#Class$ and $dim(\boldsymbol{x}) \gg \#Class$.
For GaAN, since it required the dimensions of the input/output feature vectors must be the same, $d^0_{in}=d^0_{out}=d^1_{in}=d^1_{out}=64$ and the training time of both layers were close.

@@

In general, the most time-consuming operator in GNN calculation is still the matrix multiplication \texttt{mm} and the elementwise multiplication \texttt{mul}, \emph{making GNN training suitable for GPU}.
Although the aggregate step in the edge calculation is relatively simple (like sum and mean), the related operators \texttt{scatter\_add} and \texttt{gather} still consumed a certain amount of the time.
They have to synchronize between hardware threads to avoid updating the same aggregated vector at the same time and they conducted non-regular memory access with the access pattern determined by the edge set dynamically.
For GPU, they were less efficient than \texttt{mm}.
The index-based selection \texttt{index\_select} operator from the collection step consumed around 10\% of the time in all GNNs.
Though GPUs have high on-chip memory bandwidth, improving the efficiency of \texttt{scatter\_add}/\texttt{gather}/\texttt{index\_select} (like overlapping them with $\phi$) can benefit the training of all kinds of GNNs.

### 2020-08-21 Manuscript/sec6-related-work.tex

From the efficiency aspect, Yan et al. \cite{yan2020_characterizing_gcn} compare the performance characteristics of graph convolutional networks, typical graph processing (like PageRank), and MLP-based neural network on GPU.
They provide optimization guidelines for the software and the hardware sides.
Zhang et al. \cite{zhang2020_analysis_neugraph} compare the architectural characteristics of the GNN inference on GPU under the unified SAGA-NN \cite{ma2019_neugraph} programming model,
They find that GNN inference has no fixed performance bottleneck and all components deserve to optimize.
These two efforts focus on the \emph{inference} phase of GNNs and they investigate the potential optimizations from an architectural view.
In this work, our target is to find out the performance bottleneck in the training phase from a system view.
We consider the performance bottleneck in both time and memory usage.
We also evaluate the effects of the sampling techniques.
Our work and \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph} form a complementary study on the efficiency issue of GNNs.

>>>>

From the efficiency aspect, Yan et al. \cite{yan2020_characterizing_gcn} compare the performance characteristics of graph convolutional networks, typical graph processing (like PageRank), and MLP-based neural network on GPU.
They provide optimization guidelines for the software and the hardware sides.
Zhang et al. \cite{zhang2020_analysis_neugraph} compare the architectural characteristics of the GNN inference on GPU under the unified SAGA-NN \cite{ma2019_neugraph} programming model.
They find that GNN inference has no fixed performance bottleneck and all components deserve to optimize.
These two efforts focus on the \emph{inference} phase of GNNs and they investigate the potential optimizations from an architectural view.
In this work, our target is to find out the performance bottleneck in the training phase from a system view.
We consider the performance bottleneck in both time and memory usage.
We also evaluate the effects of the sampling techniques.
Our work and \cite{yan2020_characterizing_gcn, zhang2020_analysis_neugraph} form a complementary study on the efficiency issue of GNNs.
